Universal acceptance of (nearly all of) basic science in the US

### 2024-09-02 paper.tex
draft before edits Hugo

Unfortunately, trust in science is far from being at ceiling (Wellcome Global Monitor, 2018, 2020), and there are suggestions that it has recently been dropping in some countries (Algan et al., 2021) and that it is increasingly politicized in the US (Gauchat, 2012; Krause, Brossard, Scheufele, Xenos, \& Franke, 2019; Li \& Qian, 2022).

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (creationism, etc.), or beliefs that are associated with disliked behaviors (vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We'll pay special attention to two groups of participants: people who say they don't trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We presently briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

@@

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science (Wellcome Global Monitor, 2018) . In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants (Wellcome Global Monitor, 2020). Across all countries, in 2018, 32\% of participants said they trust science ``a lot'' (41\% in 2020), 45\% trust ``some'' science (39\% in 2020), and only 13\% percent trust science ``not much/ not at all'' (13\% in 2020), while 10\% indicated ``don't know'' (7\% in 2020) (\url{https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020\#gid=6acd&pid=0}).

Trust in science also appears to be relatively stable in the US (Funk \& Kennedy, 2020). However, there are some potentially worrying trends: The US might be increasingly polarized on trust in science (Gauchat, 2012). For example, the partisan divide on the question of whether climate change is a major threat to societey \href{https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/}{has been growing in the last decade} (Kennedy, 2023). And while by some measures only about half of the US population believes in anthropogenic climate change, people do mostly not seem to change their mind on on the issue (Motta, 2021).

Trust in science is typically measured in three ways: either asking explictly about general trust in science or scientists, by asking about science attitudes (e.g.~XX), or asking questions about specific, typically contentious, science topics (e.g.~vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

@@

A lot of research has been conducted on the public understanding of science. Part of this research are surveys measuring to which extent people have basic science knowledge (e.g.~``Does the Earth go around the Sun, or does the Sun go around the Earth?''). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge (Miller, 1998, 2004). For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria but not viruses (Science Literacy et al., 2016). A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

{[}To do: Make a descriptive overview of science knowledge across time/countries. This is a crucial point: if there are questions with specific scientific answers that just about everyone already knows, then in a sense our point is already proven. But there are enough questions where people's knowledge appears quite poor anyways{]}

@@

In a series of studies, we ask basic science knowledge questions, inform people of the scientific consensus, and see if they accept it. Of particular interest are people who are low in TiS, and CTists.

In study 1, we plainly inform participants of the scientific consensus and ask if they accept it. We use a subset of questions on basic science knowledge that have been used in numerous public opinion surveys (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). In study 2, we remove a problematic science question and present people with a more elaborated explanation of the scientific consensus, as well as additional sources. This allows us to remove the issue that participants might simply not have trusted us to correctly report the scientific consensus in study 1. Study 3 follows essentially the same design as study 2, but is run on a vaccine-skeptic sample. Further, we provided participants with an explicit option to revise their answer, in case they rejected the consensus. We also asked follow-up questions on why participants accepted the scientific consensus, namely whether they thought it was because they trust scientists or because they verified independently. Study 4 uses the same design as study 3, but asking science questions that people are less likely to have encountered before, and that they would be less likely to be able to verify (or even understand) themselves {[}on an anti-vaxx sample again?{]}.

Then: replications in other countries?

\section{Overview of experiments}\label{overview-of-experiments}

The goal is to see to which extent people do not accept basic science.

\section{Experiment 1}\label{experiment-1}

The main goal of experiment one was to test whether people would accept the scientific consensus on basic knowledge questions. Additionally, we wanted to know if both science knowledge and acceptance of the scientific consensus are associated with trust in science and conspiracy thinking. We had the following research questions:

\textbf{RQ1: What is the average science knowledge score (1)?}

\textbf{RQ2: What is the average acceptance of the scientific consensus (2)?}

\textbf{RQ3: What is the relationship between trust in science and, respectively, (1) and (2)?}

\textbf{RQ4: What is the relationship between conspiracy thinking and, respectively, (1) and (2)?}

\subsection{Methods}\label{methods}

\subsubsection{Participants}\label{participants}

\subsubsection{Procedure}\label{procedure}

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 11 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions, which were randomly selected from a pool of 11 questions, in random order. After each question, participants were presented with an answer reflecting the scientific consensus. Participants were asked to choose whether they accept the answer or not, before proceeding to the next question. Figure \ref{fig:stimulus-example} displays the survey for an example science question. Finally, participants answered questions on conspiracy thinking and trust in science.

\subsubsection{Materials}\label{materials}

\paragraph{Science knowledge and acceptance}\label{science-knowledge-and-acceptance}

Table \ref{tab:knowledge} shows all questions, their scientifically consensual answer, and their source. All but two questions were selected from existing science knowledge questionnaires. We tried to select non-political questions.

@@

\paragraph{Conspiracy scales}\label{conspiracy-scales}

To measure conspiracy thinking, we selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) by Pennycook, Binnendyk, and Rand (2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

@@

To cross-check our results with alternative measures, we also assessed the conspiracy mentality questionnaire (CMQ) by Bruder, Haffke, Neave, Nouripanah, and Imhoff (2013) and the Single Item Conspiracy Beliefs Scale (SICBS) by Lantian, Muller, Nurra, and Douglas (2016) (see Appendix \ref{exp1}).

\paragraph{Trust in science}\label{trust-in-science-1}

Our main item for measuring trust in science is selected from the Wellcome Global Monitor survey: ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''

We also included two additional trust questions, one also from the Wellcome Global Monitor (WGM) survey (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), the other from the Pew research center (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''). We selected these items so that we could compare the ratings in our sample to global survey results. The WGM survey has been administered in over 140 countries and included over 140000 respondents. The Pew question has recently been used by a world-wide many labs study in 67 countries with 71417 respondents (Cologna et al., 2024).

\subsection{Results}\label{results}

Are trust in science and conspiracy thinking, respectively, associated with being more easily convinced of the scientific consensus? In our main analyses, we looked at correlations of acceptance across all observations. One possibility is that the associations between trust in science/conspiracy thinking and acceptance of scientific consensus are explained by science knowledge: People who give the right answer in the first place are more ready to accept the consensus, and trust in science/conspiracy thinking are mostly associated with this knowledge, but not with willingness to accept the consensus. To addressed this potential confound, in a non-preregistered analysis, we restricted our sample to cases where participants gave the wrong answer to the knowledge question. We than calculated the correlation between trust in science and average acceptance rate by participant. We find no statistically significant correlation of acceptance with neither conspiracy thinking (r = -0.14, p = 0.061), nor with trust in science (r = 0.06, p = 0.387).

\subsection{Discussion}\label{discussion}

\section{Experiment 2}\label{experiment-2}

\subsection{Methods}\label{methods-1}

\subsubsection{Participants}\label{participants-1}

\subsubsection{Procedure}\label{procedure-1}

\subsubsection{Materials}\label{materials-1}

@@

@@

\section{Experiment 3}\label{experiment-3}

\subsection{Methods}\label{methods-2}

\subsubsection{Participants}\label{participants-2}

\subsubsection{Procedure}\label{procedure-2}

\subsubsection{Materials}\label{materials-2}

We relied on the same items as in experiment 2. We used a pre-registered categorization scheme for the open ended answers for why people reject the consensus. The categories were XX and XX. {[}Describe coding process, with independent reviewer etc.{]}

As in study 1 and 2, we find that participants answered on average 75 \% (sd = 0.18) of the questions correctly, and initially accepted the scientific consensus on average for 96 \% (sd = 0.08) of the questions (RQ2). The acceptance rate is even higher when accounting for opinion revisions towards acceptance of the consensus, after initial rejection (98 \%, sd = 0.06).

In the appendix XX we provide an analysis that this drop is statistically significant {[}TO DO{]}.

For all correlations, we include opinion revisions for measuring consensus acceptance. We find no statistically significant correlation between science knowledge and trust in science (r = 0.14, p = 0.100), but a samll positive correlation between acceptance of scientific consensus and trust in science (r = 0.20, p = 0.019). Again, these findings might be partly due to ceiling effects: As illustrated in Fig. \ref{fig:exp3-plot}, (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. We find no statistically significant correlation between conspiracy thinking and science knowledge (r = -0.16, p = 0.055), and between conspiracy thinking and acceptance of scientific consensus (r = c(cor = -0.023), c(t = -0.27), = 0.788, c(df = 136), -0.189, 0.145, Pearson's product-moment correlation, two.sided, {[}-0.189, 0.145{]}, p = 0.788).

In Appendix \ref{exp3}, we show that these associations are statistically significant (except the last one?). We show that they are robust when using alternative measures of trust and conspiracy thinking. Appendix \ref{exp3} also includes more descriptive statistics, such as knowledge and acceptance by science questions.

Regarding RQ1, participants answered on average 75 \% (sd = 0.18) of the questions correctly. For RQ3, we got 74 answers from 47 different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Table \ref{tab:exp3-justifications} summarizes these answers by five categories.

For RQ4, we had 122 participants answering the question. Of these 41.8\% said they accepted the scientific consensus because they trust scientists on this question, while 47.5\% said they independently verified the fact. 10.7\% answered with other ``other'' and gave an open-ended explanation (see Appendix).

We also asked all 58 participants who answered that they had independently verified the answer to explain how they did so. The open-ended answers are listed in Appendix XX.

\section{Discussion}\label{discussion-2}

{[}We can draw some inspiration from the ``Westwood et al 2021 Current research overstates American support for political violence'' paper: we want to put in perspective the current rise (? discourse around?) in mistrust towards science{]}

There is increasingly more talk of defiance towards science and towards experts more generally. In every population, many people only trust science `some of the time', and a sizable minority doesn't trust it. This appears difficult to reconcile with the fact that, in many countries, most people receive at least a basic science education, that science education (along with education more generally) is increasing, and that science education seems to be the main driver of trust in science.

But how deep is distrust of science? In four studies, we have shown that almost every participant accepts almost all of basic science questions we included.

Traditionally, people have interpreted people's lack of basic knowledge in light of the deficit-model. Here, we show that people readily accept the scientifically consensual answer to these questions.

``The dominant approach to conceptualizing and measuring science literacy in population surveys has arisen out of work by Jon D. Miller and Kenneth Prewitt in the United States (see Miller, 1983, 1998, 2004) alongside collabora- tors in Great Britain (see Durant et al., 1989). Underlying these efforts appears to have been widespread concern among policy makers and the scientific com- munity that nonscientists were becoming skeptical about the benefits of sci- ence and that such skepticism might result in cuts to science funding that would harm the scientific progress that many argue underpins both American and European economic development (Bauer et al., 2007). The results of the U.S. portion of this work have formed the core of a chapter of a biennial report called Science and Engineering Indicators (hereafter, Indicators) that the National Science Board provides to Congress and the Executive Branch. Scholars have also used the raw data collected for Indicators (which is made publicly available) for peer-reviewed research (e.g., Gauchat, 2012; Losh, 2010), and other countries have used many of the Indicators' questions for their own national surveys (e.g., Bauer et al., 2012a; National Science Board, 2016).''

@@

@@

@@

@@

@@

@@

\subsubsection{Conspiracy scales}\label{conspiracy-scales-1}

@@

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-3}

@@

\subsection{Trust in science}\label{trust-in-science-4}

@@

\begin{longtable}[t]{>{}r>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:unnamed-chunk-21}Reasons for consensus rejection}\\
\toprule
id & comments\\
\midrule
1 & no\\
20 & None\\
22 & I answered thoughtfully, good luck!\\
27 & interesting, thank you.\\
30 & n/a\\
\addlinespace
32 & None\\
39 & no\\
41 & none\\
42 & n/a\\
67 & thanks\\
\addlinespace
69 & No comments\\
74 & no\\
76 & none\\
78 & None - thank you!\\
79 & Should've clicked agree about plants getting carbon from the air, but I wanted to look it up first. Whenever I encounter something I didn't know online, I like to cross-reference. In retrospect though, everything else you asked about was simple and on the level, and I'm sure you are correct.

@@

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-5}

@@

\subsection{Trust in science}\label{trust-in-science-6}

@@

\subsection{Materials}\label{materials-5}

@@

>>>>

Unfortunately, trust in science is far from being at ceiling (Wellcome Global Monitor, 2018, 2020), and there are suggestions that it has recently been dropping in some countries (Algan et al., 2021), including the US (Brian \& Tyson, 2023), where it is also increasingly politicized (Gauchat, 2012; Krause, Brossard, Scheufele, Xenos, \& Franke, 2019; Li \& Qian, 2022).

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (e.g.~creationism), or beliefs that are associated with disliked behaviors (e.g.~vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We pay special attention to two groups of participants: people who say they don't trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

@@

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science (Wellcome Global Monitor, 2018). In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants (Wellcome Global Monitor, 2020). Across all countries, in 2018, 32\% of participants said they trust science ``a lot'' (41\% in 2020), 45\% trust ``some'' science (39\% in 2020), and only 13\% percent trust science ``not much/ not at all'' (13\% in 2020), while 10\% indicated ``don't know'' (7\% in 2020)\footnote{\url{https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020\#gid=6acd&pid=0}\textgreater{}}.

Trust in science also appears to be relatively stable in the US (Funk \& Kennedy, 2020). However, there are some potentially worrying trends: Recently, trust in US has dropped (Brian \& Tyson, 2023), and some evidence suggests increasing polarization of trust in science along political lines (Gauchat, 2012). For example, the partisan divide on the question of whether climate change is a major threat to society has been growing in the last decade\footnote{\url{https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/}} (Kennedy, 2023). A recent study suggests that by some measure, only about half of the US population believes in anthropogenic climate change, and that individuals do mostly not seem to change their mind on on the issue (Motta, 2021).

Trust in science is typically measured in three ways: either asking explicitly about general trust in science or scientists, by asking about science attitudes (e.g.~support for public funding of science), or asking questions about specific, typically contentious, science topics (e.g.~vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

@@

Much of the research on public understanding of science measures to which extent people have basic science knowledge (e.g.~``Does the Earth go around the Sun, or does the Sun go around the Earth?''). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge (Miller, 1998, 2004). For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses (Science Literacy et al., 2016). A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

{[}To do: Make a descriptive overview of science knowledge across time/countries.{]}

@@

In a series of four pre-registered studies (n = 782), we asked participants about basic (studies 1, 2 and 3) and recent/advanced (study 4) science knowledge questions, informed them of the scientific consensus, and aked if they accept it. Of particular interest were people who are low in TiS, and CTists. Our main hypotheses were:

\textbf{H1: Higher trust in science is associated with more science knowledge and more acceptance of the scientific consensus}

\textbf{H2: Higher conspiracy thinking/belief is associated with less science knowledge and less acceptance of the scientific consensus}

In study 1, we plainly inform participants of the scientific consensus and ask if they accept it. We use a subset of questions on basic science knowledge that have been used in numerous public opinion surveys (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). In study 2, we remove a problematic science question and present people with a more elaborated explanation of the scientific consensus, as well as additional sources. This allows us to remove the issue that participants might simply not have trusted us to correctly report the scientific consensus in study 1. Study 3 follows essentially the same design as study 2, but is run on a vaccine-skeptic sample. Further, we provided participants with an explicit option to revise their answer, in case they rejected the consensus. We also asked follow-up questions on why participants accepted the scientific consensus, namely whether they thought it was because they trust scientists or because they verified independently. Study 4 uses the same design as study 3, but using a set of more recent, advance science questions that participants are less likely to have encountered before, and that they would be less likely to be able to verify (or even understand) themselves. Just as study 3, study 4 is run on a sample of vaccine-skeptic participants.

\section{Methods}\label{methods}

\subsection{Procedure}\label{procedure}

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions (in study 1 they were randomly selected from a pool of 11 questions) in random order. After each question, participants were presented with an answer reflecting the scientific consensus. In study 2 and 3, participants additionally saw a short explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In study 4, we provided only two links and no explanation. Participants were asked to choose whether they accept the answer or not, before proceeding to the next question. Participants then answered questions on conspiracy thinking, conspiracy beliefs and trust in science.

In study 2, 3 and 4, we presented participants with open-ended questions to explain cases where they rejected the scientific consensus. In study 3 and 4, we additionally gave participants the option to change their answer and accept the scientific consensus. Finally, at the end of study 3 and 4, we asked participants: ``For the questions in which you agreed with the scientific consensus, would you say that\ldots?'' The answer options were: (i) ``You mostly agree with the consensus because, on that question, you trust scientists'', (ii) ``You mostly agree with the consensus because you have been able to independently verify it'', and (iii) ``Other'', with a text box for participants to explain. Participants who selected ``You mostly agree with the consensus because you have been able to independently verify it'', were asked the open-ended follow-up question: ``Could you please tell us how you independently verified the information?''.

\subsection{Participants}\label{participants}

We recruited a total of 782 (194 in study 1, 190 in study 2, 200 in study 3, 198 in study 4) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for study 1 and 2 were convenience samples, study 3 and 4 were run on a sample holding vaccine-skeptic beliefs (see OSM for details).

\subsection{Materials}\label{materials}

Table \ref{tab:knowledge} shows all questions and their answer options used in the four studies.

@@

Our main measure of conspiracy belief was asking participants about specific conspiracy theories. We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) (Pennycook, Binnendyk, \& Rand, 2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

As a robustness check, we also used established scales on conspiracy thinking, namely the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the OSM.

@@

Our main item for measuring trust in science was selected from the Wellcome Global Monitor survey: ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''. We also included two additional trust questions, one also from the Wellcome Global Monitor (WGM) survey (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), the other from the Pew research center (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''). We selected these items so that we could compare the ratings in our sample to global survey results. The WGM survey has been administered in over 140 countries and included over 140000 respondents. The Pew question has recently been used by a world-wide many labs study in 67 countries with 71417 respondents (Cologna et al., 2024).

\section{Results}\label{results}

Across all studies we find that almost all participants trust basic and recent/advance science facts almost all the time. That is true even for participants who expressed low general trust in science, and those who have a tendency to conspiracy think or hold science-related conspiracy beliefs (Fig. \ref{fig:summary-plot}).

Our results suggest that the more people tend to trust science, the more they tend to know about science and accept the scientific consensus. In study 2 and 3, we find a positive but small correlation between trust in science and both science knowledge (study 1: r = 0.29, p \textless{} .001; study 2: r = 0.28, p \textless{} .001), and acceptance of scientific consensus (study 1: r = 0.27, p \textless{} .001; study 2: r = 0.30, p \textless{} .001). In study 3 and 4--both run on vaccine-skeptic samples--we find no statistically significant correlation between science knowledge and trust in science (study 3: r = 0.14, p = 0.100; study 4: r = 0.10, p = 0.144), but a small positive correlation between acceptance of scientific consensus and trust in science (study 3: r = 0.20, p = 0.019; study 4: r = 0.16, p = 0.029).

As for conspiracy beliefs, results are mixed. In studies 1 and 2, we find a small negative correlation between conspiracy thinking and both science knowledge (study 1: r = -0.38, p \textless{} .001; study 2: r = -0.40, p \textless{} .001), and acceptance of scientific consensus (study 1: r = -0.33, p \textless{} .001; study 2: r = -0.37, p \textless{} .001). However, in study 3 and 4, we do not find a statistically significant correlation with neither science knowledge (study 3: r = -0.16, p = 0.055; study 4: r = -0.02, p = 0.742), and between nor acceptance of scientific consensus (study 3: r = -0.02, p = 0.788; study 4: r = -0.05, p = 0.498).

Overall, the correlations we find are relatively weak, which might be partly due to ceiling effects. This was the case in particular for acceptance: In all studies, in the vast majority of cases, participants accepted accepted science facts. The absence of correlations between conspiracy beliefs and science knowledge or acceptance in studies 3 and 4 might be explained by our sample selection. For these studies, we recruited samples scoring considerably higher in conspiracy belief than the convenience samples of study 1 and 2 (see OSM). The presence of effects in study 1 and 2, and the absence in study 3 and 4 might be because differences between moderate and high conspiracy thinking might not matter, while, differences between low and moderate/high conspiracy thinking do matter.

Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. Only in study 4 do we find evidence\footnote{For this analysis, we restricted our data to those cases in which participants initially gave the wrong answer.} that changing one's mind towards the scientific consensus is associated with (more) trust in science (study 1: r = 0.06, p = 0.387; study 2: r = 0.16, p = 0.051; study 3: r = 0.04, p = 0.619; study 4: r = 0.15, p = 0.037) and only in study 2 evidence that it is associated with (less) conspiracy thinking (study 1: r = -0.14, p = 0.061; study 2: r = -0.22, p = 0.006; study 3: r = -0.04, p = 0.631; study 4: r = -0.05, p = 0.455). In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: 1.6 \%; study 2: 0.5 \%; study 3: 0.7 \%; study 4: 5.7 \%).

Why did participants reject the scientific consensus? We got a total of 364 answers (study 2: 35; study 3: 74; study 4: 255) from 167 (study 2: 25; study 3: 47; study 4: 95) different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories. Table \ref{tab:justifications} summarizes these answers by five categories. All answers can be read in the OSM.

@@

\caption{\label{tab:justifications}Justifications by category for data from studies 2, 3 and 4 combined.}

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
NA & 255.00 & 70.1\% & 95.00\\
Personal convictions & 37.00 & 10.2\% & 27.00\\
Mistake & 30.00 & 8.2\% & 26.00\\
No justification & 25.00 & 6.9\% & 18.00\\
Not convinced & 12.00 & 3.3\% & 8.00\\
Religious Beliefs & 5.00 & 1.4\% & 5.00\\
\bottomrule
\end{tabular}

@@

Why did participants accept the scientific consensus? In studies 3 and 4, we had asked participants about cases where they agreed with the scientific consensus. A total of 320 (study 3: 122; study 4: 198) participants answered this question. Of these, 41.8\% in study 3 and 36.4\% in study 4 said they accepted the scientific consensus because they trust scientists, while 47.5\% in study 3 and 47\% in study 4 said they independently verified the fact. We asked all participants who gave this answer to explain how they did it (answers can be found in the OSM). 10.7\% in study 3 and 16.7\% in study 4 answered with other ``other'' and gave an open-ended explanation (see OSM).

More detailed results addressing all our pre-registered research questions can be found in the OSM.

\section{Discussion}\label{discussion}

Survey data from the US suggests a decline of trust in science in recent years (Brian \& Tyson, 2023). Across the globe, many people only trust science `some of the time' , and a sizable minority doesn't trust it (Wellcome Global Monitor, 2018, 2020). This appears difficult to reconcile with the fact that, in many countries, most people nowadays receive at least a basic education (Ritchie, Samborska, Ahuja, Ortiz-Ospina, \& Roser, 2023), and that science education has been found to be the most strongly associated variable with trust in science (Wellcome Global Monitor, 2018).

But how deep is distrust of science? In four studies, we have shown that almost every participant accepts almost all of basic science questions we included. Even when people do not know the correct answer to a science question, they tend to mostly accept the scientific consensus afterwards. That is true not only for basic science questions (studies 1-3), but also for more recent scientific findings that most people are unfamiliar with (study 4). In study 1 the acceptance rate after having given a wrong answer was considerably lower (23.7 \%) than in subsequent studies (study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%). While this could be just sampling variation, it might be that adding explanations and/or sources convinced participants more than merely stating the consensus. In general, participants with lower trust in science and who believe more in conspiracy theories tended to both know less about science and accept the scientific consensus less.

Scholars have long interpreted the lack of basic science knowledge as the root cause of distrust in and more generally unfavorable attitudes towards science, an explanation known under the term ``deficit-model'' (Durant et al., 1989; Miller, 1998, 2004). In line with proponents of the deficit-model, we find that people indeed err even on very basic science questions. However, we also find that almost everyone instantly accepted the scientific consensus once informed about it. Moreover, the same was true for non-basic science questions. This calls into question the deficit model: If participants trust science on basic (and non-basic) questions---as suggests immediate acceptance of the scientific consensus---then unfavorable attitudes towards science must have other explanations than a mere lack of knowledge. This argument is supported by past research. In a seminal meta-analysis, @allumScienceKnowledgeAttitudes2008 found correlations between science knowledge and general attitudes towards science to be at best small and largely context dependent {[}@allumScienceKnowledgeAttitudes2008{]}. A more recent report by the National Academies of Sciences, Engineering, and Medicine states that science knowledge is ``unlikely to substantially affect attitudes on scientific issues'' (National Academies of Sciences, 2017).

How to make sense of the detachment of basic science knowledge from attitudes towards science? One explanation might be that most scientific findings are very abstract. For example, to most people, it will not be relevant for their daily lives to know that the speed of light is 299'792'458 meters per second. It is therefor easy to simply accept it, all the more if it appears to be consensual within the population of scientists. In other words, people might hold on to most scientific findings as reflective beliefs---beliefs that have little effect on our behaviors (Mercier, 2021). It is probably when it becomes tangible that science shape's people's attitudes. This can happen, for example, when science appears to support conclusions in favor of certain policies or moral positions, as for example in the case of anthropogenic climate change or vaccines.

It is tempting, especially for scientists, to explain mistrust in science with ignorance. After all, careful proof of objective truth is what makes science trustworthy---that is, for scientists. For most people, what shapes trust in science might not be so different from what shapes trust in other institutions, such as the government, banks or church.

@@

\bibitem[\citeproctext]{ref-brianAmericansTrustScientists2023}
Brian, K., \& Tyson, A. (2023). \emph{Americans{'} trust in scientists, positive views of science continue to decline}. Retrieved from \url{https://www.pewresearch.org/science/2023/11/14/americans-trust-in-scientists-positive-views-of-science-continue-to-decline/}

@@

\bibitem[\citeproctext]{ref-mercierHowGoodAre2021}
Mercier, H. (2021). How Good Are We At Evaluating Communicated Information? \emph{Royal Institute of Philosophy Supplement}, \emph{89}, 257--272. \url{https://doi.org/10.1017/S1358246121000096}

@@

\bibitem[\citeproctext]{ref-nationalacademiesofsciencesCommunicatingScienceEffectively2017}
National Academies of Sciences, E. (2017). \emph{Communicating science effectively: A research agenda}. National Academies Press.

@@

\bibitem[\citeproctext]{ref-ritchieGlobalEducation2023}
Ritchie, H., Samborska, V., Ahuja, N., Ortiz-Ospina, E., \& Roser, M. (2023). Global education. \emph{Our World in Data}. Retrieved from \url{https://ourworldindata.org/global-education}

@@

\subsection{Participants}\label{participants-1}

\subsection{Materials}\label{materials-1}

@@

\subsubsection{Conspiracy scales}\label{conspiracy-scales}

@@

\subsubsection{Trust in science}\label{trust-in-science-1}

@@

Are trust in science and conspiracy thinking, respectively, associated with being more easily convinced of the scientific consensus? In our main analyses, we looked at correlations of acceptance across all observations. One possibility is that the associations between trust in science/conspiracy thinking and acceptance of scientific consensus are explained by science knowledge: People who give the right answer in the first place are more ready to accept the consensus, and trust in science/conspiracy thinking are mostly associated with this knowledge, but not with willingness to accept the consensus. To addressed this potential confound, in a non-preregistered analysis, we restricted our sample to cases where participants gave the wrong answer to the knowledge question. We then calculated the correlation between trust in science and average acceptance rate by participant. We find no statistically significant correlation of acceptance with neither conspiracy thinking (r = -0.14, p = 0.061), nor with trust in science (r = 0.06, p = 0.387).

@@

@@

\subsection{Trust in science}\label{trust-in-science-3}

@@

\begin{longtable}[t]{>{}r>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:unnamed-chunk-22}Reasons for consensus rejection}\\
\toprule
id & comments\\
\midrule
1 & no\\
20 & None\\
22 & I answered thoughtfully, good luck!\\
27 & interesting, thank you.\\
30 & n/a\\
\addlinespace
32 & None\\
39 & no\\
41 & none\\
42 & n/a\\
67 & thanks\\
\addlinespace
69 & No comments\\
74 & no\\
76 & none\\
78 & None - thank you!\\
79 & Should've clicked agree about plants getting carbon from the air, but I wanted to look it up first. Whenever I encounter something I didn't know online, I like to cross-reference. In retrospect though, everything else you asked about was simple and on the level, and I'm sure you are correct.

@@

\subsection{Participants}\label{participants-2}

\subsection{Procedure}\label{procedure-1}

\subsection{Materials}\label{materials-2}

@@

\subsection{Discussion}\label{discussion-2}

@@

\subsubsection{Trust in science}\label{trust-in-science-4}

@@

\subsection{Trust in science}\label{trust-in-science-5}

@@

\subsection{Participants}\label{participants-3}

\subsection{Procedure}\label{procedure-2}

@@

\subsection{Results}\label{results-3}

As in study 1 and 2, we find that participants answered on average 75 \% (sd = 0.18) of the questions correctly (RQ1), and initially accepted the scientific consensus on average for 96 \% (sd = 0.08) of the questions (RQ2). The acceptance rate is even higher when accounting for opinion revisions towards acceptance of the consensus, after initial rejection (98 \%, sd = 0.06).
